Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-05-19 article-models-ttinternal.tex
reorganizing the weakening section

\begin{defn}
An \emph{extension-empty algebra in context $\Gamma$} 
is a sextuple $(A,P,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ for which 
the quadruple $(A,P,\epsilon_0,\epsilon_1)$ is an extension algebra in context 
$\Gamma$, the quadruple $(A,P,\phi_0,\phi_1)$ is an empty algebra in context
$\Gamma$, satisfying the judgmental equalities 
\autoref{extempalg-eq1,extempalg-eq2,extempalg-eq3,extempalg-eq4}.
\end{defn}

@@

To complete the proof, we need to verify that
\autoref{extempalg-eq1,extempalg-eq2,extempalg-eq3,extempalg-eq4} hold in our
current setting. More precisely, we have to verify that
\begin{align*}
\subst{\phi_1}{\epsilon_1}
& \jdeq 
  \idtm{P}
  \tag{1}
  \\
\subst{\jcomp{}{\epsilon_0}{\phi_1}}{\epsilon_1}
& \jdeq 
  \idtm{P}
  \tag{2}
  \\
\subst{\phi_1}{\jcomp{}{\epsilon_0}{\epsilon_1}}
& \jdeq
  \epsilon_1
  \tag{3}
  \\
\subst{\jcomp{}{\epsilon_0}{\phi_1}}{\jcomp{}{\epsilon_0}{\epsilon_1}}
& \jdeq
  \idtm{\jcomp{}{\epsilon_0}{P}}
  \tag{4}
  \\
\end{align*}
\end{proof}

@@

\begin{thm}
Let $\mathcal{A}\defeq(A,P,T,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ be a term-algebra in context
$\Gamma$. Then the septuple
\begin{equation*}
(P,\jcomp{}{\epsilon_0}{P},\jcomp{}{\epsilon_0}{T},\epsilon_1,\jcomp{}{\epsilon_0}{\epsilon_1},\phi_1,\jcomp{}{\epsilon_0}{\phi_1})
\end{equation*}
is a term-algebra in context $\ctxext{\Gamma}{A}$, called the family term-algebra
of $\mathcal{A}$.
\end{thm}

@@

Let $(A,P,T,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ be a term-algebra in context
$\Gamma$.

An extension-weakening algebra in context $\Gamma$ will be an octuple
\begin{equation*}
(A,P,\epsilon_0,\epsilon_1,\phi_0,\phi_1,\omega_0,\omega_1)
\end{equation*}
where $(A,P,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ is an extension-empty algebra
in context $\Gamma$ and where
\begin{align*}
\jhom*
  {{{\Gamma}{A}}{P}}
  {\ctxwk{P}{P}}
  {\jcomp{}{\epsilon_0}{P}}
  {\omega_0}
  \\
\jfhom*
  {{{\Gamma}{A}}{P}}
  {\ctxwk{P}{P}}
  {\jcomp{}{\epsilon_0}{P}}
  {\omega_0}
  {\ctxwk{P}{\jcomp{}{\epsilon_0}{P}}}
  {\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}}}
  {\omega_1}
\end{align*}
such that $(\omega_0,\omega_1)$ is both an extension homomorphism
\begin{equation*}
\begin{tikzcd}[column sep=large]
\ctxwk{P}{(P,
  \jcomp{}{\epsilon_0}{P},
  \epsilon_1,
  \jcomp{}{\epsilon_0}{\epsilon_1})}
\ar{r}{(\omega_0,\omega_1)}
& \jcomp{}{\epsilon_0}{(P,
  \jcomp{}{\epsilon_0}{P},
  \epsilon_1,
  \jcomp{}{\epsilon_0}{\epsilon_1})}
\end{tikzcd}
\end{equation*}
and an empty homomorphism

satisfying five(?) additional judgmental equalities expressing that $\omega_0$
and $\omega_1$ are compatible with $\epsilon_1$, $\phi_1$
and with each other. Our current goal is to figure out what these are.

Before we go into that, we develop a bit of intuition by explaining how
$\omega_0$ and $\omega_1$ act when applied to a family $x_1:\subst{x_0}{P}$
over $x_0:A$. Note that
\begin{equation*}
\jterm{{\Gamma}{\subst{x_0}{P}}}{\subst{{x_1}{{x_0}{\epsilon_0}}}{P}}{\subst{x_1}{{x_0}{\omega_0}}},
\end{equation*}
so $\subst{x_1}{{x_0}{\omega_0}}$ takes families over $x_0$ to families over
the extended $\subst{x_1}{{x_0}{\epsilon_0}}$.

@@

\subsubsection{The compatibility of weakening with extension}
The first rule, expressing that $\omega_0$ is compatible with $\epsilon_1$ is that
the diagram
\begin{equation}\label{wkalg-exteq1}
\begin{tikzcd}[column sep=large]
\ctxwk{P}{\ctxext{P}{\jcomp{}{\epsilon_0}{P}}}
  \ar{r}{\jvcomp{}{\omega_0}{\omega_1}}
  \ar{d}[swap]{\ctxwk{P}{\epsilon_1}}
& \ctxext
    {\jcomp{}{\epsilon_0}{P}}
    {\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}}}
  \ar{d}{\jcomp{}{\epsilon_0}{\epsilon_1}}
  \\
\ctxwk{P}{P}
  \ar{r}[swap]{\omega_0}
& \jcomp{}{\epsilon_0}{P}
\end{tikzcd}
\end{equation}
commutes judgmentally. For the second rule, expressing that $\omega_1$ is 
compatible with $\epsilon_1$, we wish to fill in a commutative diagram
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\ctxwk
  {P}
  { \ctxext
      {\jcomp{}{\epsilon_0}{P}}
      {\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}}}
    }
  \ar[densely dotted]{r}{\jvcomp{}{\omega_1}{?}}
  \ar{d}[swap]{\ctxwk{P}{\jcomp{}{\epsilon_0}{\epsilon_1}}}
& \jcomp{}{\omega_0}{%
    \jcomp{}{\epsilon_0}{%
      \jcomp{}{\epsilon_0}{%
        \ctxext{P}{\jcomp{}{\epsilon_0}{P}}
        }
      }
    }
  \ar{d}{ \jcomp{}{\omega_0}{%
            \jcomp{}{\epsilon_0}{%
              \jcomp{}{\epsilon_0}{\epsilon_1}
              }
            }
          }
  \\
\ctxwk{P}{\jcomp{}{\epsilon_0}{P}}
  \ar{r}[swap]{\omega_1}
& \jcomp{}{\omega_0}{%
    \jcomp{}{\epsilon_0}{%
      \jcomp{}{\epsilon_0}{P}
      }
    }
\end{tikzcd}
\end{equation*}
The morphism indicated by the question mark is a morphism from
$\ctxwk{P}{\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}}}$ to
$%
\jcomp{}{\omega_1}{%
  \jcomp{}{\omega_0}{%
    \jcomp{}{\epsilon_0}{%
      \jcomp{}{\epsilon_0}{%
        \ctxext{P}{\jcomp{}{\epsilon_0}{P}}
        }
      }
    }
  }
$ in context $\ctxext{{\Gamma}{A}}{P}$. One would expect this morphism to be
\begin{equation*}
\begin{tikzcd}[column sep=large]
\jcomp{}{\ctxwk{P}{\epsilon_0}}{\ctxwk{P}{\jcomp{}{\epsilon_0}{P}}}
  \ar{r}{\jcomp{}{\ctxwk{P}{\epsilon_0}}{\omega_1}}
& \jcomp{}{\ctxwk{P}{\epsilon_0}}{%
    \jcomp{}{\omega_0}{%
      \jcomp{}{\epsilon_0}{%
        \jcomp{}{\epsilon_0}{%
          P
          }
        }
      }
    }
\end{tikzcd}
\end{equation*}
in the context $\ctxext{{\Gamma}{A}}{P}$, 
so we must verify the judgmental equalities
\begin{align*}
\jcomp{}{\ctxwk{P}{\epsilon_0}}{\ctxwk{P}{\jcomp{}{\epsilon_0}{P}}}
& \jdeq 
  \ctxwk{P}{\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}}}
  \\
\jcomp{}{\ctxwk{P}{\epsilon_0}}{%
    \jcomp{}{\omega_0}{%
      \jcomp{}{\epsilon_0}{%
        \jcomp{}{\epsilon_0}{%
          P
          }
        }
      }
    }
& \jdeq
  \jcomp{}{\omega_1}{%
    \jcomp{}{\omega_0}{%
      \jcomp{}{\epsilon_0}{%
        \jcomp{}{\epsilon_0}{%
          \ctxext{P}{\jcomp{}{\epsilon_0}{P}}
          }
        }
      }
    } 
\end{align*}
of families over the context $\ctxext{{\Gamma}{A}}{P}$. The first of these
judgmental equalities is nothing special: it follows at straight from
\autoref{comp-ws-f}. We prove the second equality in the following lemma.

@@

This enables us to require that the diagram
\begin{equation}\label{wkalg-exteq2}
\begin{tikzcd}[column sep=huge]
\ctxwk
  {P}
  { \ctxext
      {\jcomp{}{\epsilon_0}{P}}
      {\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}}}
    }
  \ar{r}{\jvcomp{}{\omega_1}{\jcomp{}{\ctxwk{P}{\epsilon_0}}{\omega_1}}}
  \ar{d}[swap]{\ctxwk{P}{\jcomp{}{\epsilon_0}{\epsilon_1}}}
& \jcomp{}{\omega_0}{%
    \jcomp{}{\epsilon_0}{%
      \jcomp{}{\epsilon_0}{%
        \ctxext{P}{\jcomp{}{\epsilon_0}{P}}
        }
      }
    }
  \ar{d}{ \jcomp{}{\omega_0}{%
            \jcomp{}{\epsilon_0}{%
              \jcomp{}{\epsilon_0}{\epsilon_1}
              }
            }
          }
  \\
\ctxwk{P}{\jcomp{}{\epsilon_0}{P}}
  \ar{r}[swap]{\omega_1}
& \jcomp{}{\omega_0}{%
    \jcomp{}{\epsilon_0}{%
      \jcomp{}{\epsilon_0}{P}
      }
    }
\end{tikzcd}
\end{equation}
commutes judgmentally.

>>>>

\begin{defn}
An \emph{extension-empty algebra in context $\Gamma$} 
is a sextuple $(A,P,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ for which 
the quadruple $(A,P,\epsilon_0,\epsilon_1)$ is an extension algebra in context 
$\Gamma$, the quadruple $(A,P,\phi_0,\phi_1)$ is an empty algebra in context
$\Gamma$, satisfying the judgmental equalities 
\autoref{extempalg-eq1,extempalg-eq2,extempalg-eq3,extempalg-eq4,extempalg-eq5}.
\end{defn}

@@

To complete the proof, we need to verify that
\autoref{extempalg-eq1,extempalg-eq2,extempalg-eq3,extempalg-eq4,extempalg-eq5} hold in our
current setting. More precisely, we have to verify that
\begin{align*}
\subst{\phi_1}{\epsilon_1}
& \jdeq 
  \idtm{P}
  \\
\subst{\jcomp{}{\epsilon_0}{\phi_1}}{\epsilon_1}
& \jdeq 
  \idtm{P}
  \\
\subst{\phi_1}{\jcomp{}{\epsilon_0}{\epsilon_1}}
& \jdeq
  \epsilon_1
  \\
\subst{\jcomp{}{\epsilon_0}{\phi_1}}{\jcomp{}{\epsilon_0}{\epsilon_1}}
& \jdeq
  \idtm{\jcomp{}{\epsilon_0}{P}}
  \\
\subst{\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{\phi_1}}}{\jcomp{}{\epsilon_0}{\epsilon_1}}
& \jdeq
  \idtm{\jcomp{}{\epsilon_0}{P}}
\end{align*}
None of these present any difficulties.
\end{proof}

@@

\begin{thm}
Let $\mathcal{A}\defeq(A,P,T,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ be a term-algebra in context
$\Gamma$. Then the septuple
\begin{equation*}
\mathcal{F}_{\mathcal{A}}\defeq(P,\jcomp{}{\epsilon_0}{P},\jcomp{}{\epsilon_0}{T},\epsilon_1,\jcomp{}{\epsilon_0}{\epsilon_1},\phi_1,\jcomp{}{\epsilon_0}{\phi_1})
\end{equation*}
is a term-algebra in context $\ctxext{\Gamma}{A}$, called the family term-algebra
of $\mathcal{A}$.
\end{thm}

@@

Let $(A,P,T,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ be a term-algebra in context
$\Gamma$. The weakening homomorphism $\omega$ will be a term-homomorphism from 
$\ctxwk{P}{\mathcal{F}_\mathcal{A}}$ to 
$\jcomp{}{\epsilon_0}{\mathcal{F}_\mathcal{A}}$ satisfying certain extra
judgmental equalities. The goal of this subsection is to spell these out.

Concretely, the weakening homomorphism is a triple $(\omega_0,\omega_1,\omega_2)$
consisting of
\begin{align*}
\jhom*
  {{{\Gamma}{A}}{P}}
  {\ctxwk{P}{P}}
  {\jcomp{}{\epsilon_0}{P}}
  {\omega_0}
  \\
\jfhom*
  {{{\Gamma}{A}}{P}}
  {\ctxwk{P}{P}}
  {\jcomp{}{\epsilon_0}{P}}
  {\omega_0}
  {\ctxwk{P}{\jcomp{}{\epsilon_0}{P}}}
  {\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}}}
  {\omega_1}
  \\
\jfhom*
  {{{\Gamma}{A}}{P}}
  {\ctxwk{P}{\ctxext{P}{\jcomp{}{\epsilon_0}{P}}}}
  {\jcomp{}{\epsilon_0}{\ctxext{P}{\jcomp{}{\epsilon_0}{P}}}}
  {\jvcomp{\ctxwk{P}{\jcomp{}{\epsilon_0}{P}}}{\omega_0}{\omega_1}}
  {\ctxwk{P}{\jcomp{}{\epsilon_0}{T}}}
  {\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{T}}}
  {\omega_2}
\end{align*}
We could also summarize the situation in a diagram:
\begin{equation*}
\begin{tikzcd}
\ctxwk{P}{\jcomp{}{\epsilon_0}{T}}
  \ar[fib]{d}
  \ar{r}{\omega_2}
& \jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{T}}
  \ar[fib]{d}
  \\
\ctxwk{P}{\jcomp{}{\epsilon_0}{P}}
  \ar[fib]{d}
  \ar{r}{\omega_1}
& \jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}}
  \ar[fib]{d}
  \\
\ctxwk{P}{P}
  \ar{r}[swap]{\omega_0}
& \jcomp{}{\epsilon_0}{P}
\end{tikzcd}
\end{equation*}

Before we go into chasing the judgmental equalities that are to be imposed on
weakening, we develop a bit of intuition by explaining how
$\omega_0$ and $\omega_1$ act when applied to a family $x_1:\subst{x_0}{P}$
over $x_0:A$, and we will provide the explicit requirements that make
$\omega$ a term-homomorphism. Note that
\begin{equation*}
\jterm{{\Gamma}{\subst{x_0}{P}}}{\subst{{x_1}{{x_0}{\epsilon_0}}}{P}}{\subst{x_1}{{x_0}{\omega_0}}},
\end{equation*}
so $\subst{x_1}{{x_0}{\omega_0}}$ takes families over $x_0$ to families over
the extended $\subst{x_1}{{x_0}{\epsilon_0}}$.

@@

Because the pair $(\omega_0,\omega_1)$ is required to be an extension homomorphism,
the diagrams
\begin{equation}\label{wkalg-exteq1}
\begin{tikzcd}[column sep=large]
\ctxwk{P}{\ctxext{P}{\jcomp{}{\epsilon_0}{P}}}
  \ar{r}{\jvcomp{}{\omega_0}{\omega_1}}
  \ar{d}[swap]{\ctxwk{P}{\epsilon_1}}
& \ctxext
    {\jcomp{}{\epsilon_0}{P}}
    {\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}}}
  \ar{d}{\jcomp{}{\epsilon_0}{\epsilon_1}}
  \\
\ctxwk{P}{P}
  \ar{r}[swap]{\omega_0}
& \jcomp{}{\epsilon_0}{P}
\end{tikzcd}
\end{equation}
in context $\ctxext{{\Gamma}{A}}{P}$, and
\begin{equation}\label{wkalg-exteq2}
\begin{tikzcd}[column sep=huge]
\ctxwk
  {P}
  { \ctxext
      {\jcomp{}{\epsilon_0}{P}}
      {\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}}}
    }
  \ar{r}{\jvcomp{}{\omega_1}{\jcomp{}{\ctxwk{P}{\epsilon_1}}{\omega_1}}}
  \ar{d}[swap]{\ctxwk{P}{\jcomp{}{\epsilon_0}{\epsilon_1}}}
& \jcomp{}{\omega_0}{%
    \jcomp{}{\epsilon_0}{%
      \jcomp{}{\epsilon_0}{%
        \ctxext{P}{\jcomp{}{\epsilon_0}{P}}
        }
      }
    }
  \ar{d}{ \jcomp{}{\omega_0}{%
            \jcomp{}{\epsilon_0}{%
              \jcomp{}{\epsilon_0}{\epsilon_1}
              }
            }
          }
  \\
\ctxwk{P}{\jcomp{}{\epsilon_0}{P}}
  \ar{r}[swap]{\omega_1}
& \jcomp{}{\omega_0}{%
    \jcomp{}{\epsilon_0}{%
      \jcomp{}{\epsilon_0}{P}
      }
    }
\end{tikzcd}
\end{equation}
in context $\ctxext{{{\Gamma}{A}}{P}}{\ctxwk{P}{P}}$, commute judgmentally.
To see that the morphism $\jvcomp{}{\omega_1}{\jcomp{}{\ctxwk{P}{\epsilon_0}}{\omega_1}}$
indeed has the suggested codomain, we have the following lemma:

@@
